---
title: "Identifying Areas of Potential Aquaculture in the Western US"
format: pdf
editor: visual
---

**Author** William Mullins \| Date: December 10, 2025 [View Repository on Github](https://github.com/willrmull/potential_aquaculture.git)

## Introduction

In this analysis, we identified areas along the west coast of the United States which contained suitable depth and sea surface temperature (SST) for aquaculture, specifically on two species: oysters and blue mussels.

------------------------------------------------------------------------

### Datasets Used

-   [NOAA Pathfinder Sea Surface Temperature Data](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php): Data from this is used to find the five-year average of sea surface temperature data from 2008-2012, providing a baseline for temperature analysis along the western US coast.
-   [NOAA Bathymetry Data](https://www.gebco.net/data-products/gridded-bathymetry-data#area): High-resolution ocean depth measurements used to identify suitable depths for shellfish cultivation.
-   [West Coast Regional Boundaries](https://www.marineregions.org/eez.php): Spatial boundaries defining the Exclusive Economic Zones (EEZs) for California, Oregon, and Washington, used to calculate suitable habitat area by region.

------------------------------------------------------------------------

### Loading Packages

```{r, warning = FALSE, message = FALSE}
reqired_packages <- c("terra", "tidyverse", "here", "knitr", "kableExtra")

for(pkg in reqired_packages){
  if (!require(pkg, character.only = TRUE)){
     stop(paste0("Package '", pkg, "' is not installed"))
  } 
}

# Load required packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(terra)
  library(here)
  library(knitr)
  library(kableExtra)
})
```

------------------------------------------------------------------------

### Data Import and Processing

------------------------------------------------------------------------

### Data Import

The following section imports the three datasets used in the analysis. We implement validation checks to ensure files exist before attempting to load them.

-   The SST data are stored as individual GeoTIFF files, each representing the average annual SST for a given year. These rasters are stacked into a single multi-layer raster object.
-   The bathymetry raster is then imported and reprojected to match the coordinate reference system of the SST stack.
-   The EEZ boundaries are read as a vector layer.

```{r, warning = FALSE, message = FALSE}
sst <- rast(c(here("data", "average_annual_sst_2008.tif"),
              here("data", "average_annual_sst_2009.tif"),
              here("data", "average_annual_sst_2010.tif"),
              here("data", "average_annual_sst_2011.tif"),
              here("data", "average_annual_sst_2012.tif")))

# Read in depth file
depth <- rast(here("data", "depth.tif")) %>%
  project(crs(sst))

eez_raster <- vect(here::here("data", "wc_regions_clean.shp"))
```

### Data Processing

To find the temporal mean SST across the 2008–2012 period, we compute the average across the raster stack. Since the original values are in Kelvin, we convert them to Celsius by subtracting 273.15.

```{r}
# Change Margins on top and bottom
par(oma = c(1, 0, 3, 0))

# Create an informative plot of mean SST
plot(mean_sst,
     main = "Mean Sea Surface Temperature \n(2008-2012)",
     col = hcl.colors(40, "RdYlBu", rev = TRUE),
     xlab = "Longitude",
     ylab = "Latitude",
     cex.main = 1.0,
     plg = list(title = "SST (°C)"))
```

\*Interpretation\*\*: This map reveals the characteristic temperature gradient along the US West Coast. Southern California exhibits the warmest waters (20-25°C), while the Pacific Northwest and areas influenced by coastal upwelling show cooler temperatures (10-15°C). This spatial variation in temperature has important implications for species-specific aquaculture siting, as different shellfish species have distinct thermal tolerances.

### Depth Data Preparation

The depth raster is resampled to match the resolution and extent of the SST data using nearest neighbor interpolation. This ensures that both datasets align perfectly for subsequent analysis.

```{r, warning = FALSE, message = FALSE}
# Resample depth using nearest neighbor approach
depth <- resample(depth, mean_sst, method = "near")

# Crop raster to match SST
depth_cropped<- crop(depth, mean_sst)

# check that the depth and SST match in resolution, extent, and coordinate reference system
compareGeom(depth_cropped, mean_sst, res = TRUE, ext = TRUE, crs = TRUE)

# Verify alignment of datasets
alignment_check <- compareGeom(depth_cropped, mean_sst, res = TRUE, ext = TRUE, crs = TRUE)


if (!alignment_check) {
  stop(
    "Depth and SST rasters are not properly aligned! Check resolution, extent, and CRS of both datasets. ",
    call. = FALSE
  )
} else {
  message("Depth and SST rasters are properly aligned (resolution, extent, and CRS match)")
}
```

## **Habitat Suitability Analysis**

To evaluate the potential of locations for marine aquaculture of a given species, we use measurements of environmental tolerances. Using these thresholds, each pixel in the study area is classified as either suitable or unsuitable based on whether it simultaneously satisfies both temperature and depth criteria.

\### Suitability Function

The \`find_suitable_habitat()\` function implements this logic by first generating binary masks for temperature and depth suitability. The rasters are then combined, producing a categorical raster containing locations where both conditions are met.

```{r, warning = FALSE, message = FALSE}
# Function for identifying and plotting suitable habitats
find_suitable_habitat <- function(temp_min, temp_max, depth_min, depth_max, species_name) {
  
  # Create binary masks for suitable conditions
  sst_suitable <- mean_sst >= temp_min & mean_sst <= temp_max
  depth_suitable <- depth_cropped >= depth_min & depth_cropped <= depth_max
  
  # Combine masks where both conditions are met
  suitable_habitat <- sst_suitable * depth_suitable
  
  # Convert to factor
  suitable_habitat <- as.factor(suitable_habitat)
  levels(suitable_habitat) <- c("Unsuitable", "Suitable")
  
  # Create informative plot
  par(oma = c(1, 1, 1.5, 1))
  
  plot(suitable_habitat,
       col = c("#D3D3D3", "#228B22"),
       type = "classes",
       main = paste0("Potential ", species_name, " Aquaculture Habitat\nSST: ", temp_min, "–", temp_max, "°C | Depth: ", depth_min, "–", depth_max, " m"),
       xlab = "Longitude",
       ylab = "Latitude",
       axes = TRUE,
       cex.main = 0.8,
       plg = list(x = "topright", title = "Suitability"))
  
  # Return the raster
  return(suitable_habitat)
}
```

### Map of Potential Oyster Habitats

Oysters prefer slighly warm waters (11-30°C) and are able to be cultivated in shallow depths (0-70 meters).

```{r, warning = FALSE, message = FALSE}
# View Areas Habitable by Oysters
oyster_habitat <- find_suitable_habitat(
  temp_min = 11,
  temp_max = 30,
  depth_min = 0,
  depth_max = 70,
  species_name = "Oyster"
)
```

**Interpretation**: This map shows that suitable areas are concentrated along the coast of Southern California and in Washington. Cincentration is low along the coast of Northern California and Oregon. The cooler waters of Oregon and Washington limit suitable habitat primarily to nearshore areas where depths fall within the required range.

### Map of Potential Mussel Habitats

The species of blue mussels native to the West Coast, *Mytilus trossulus*, are much more cold-adapted in comparison to oysters. Although estimates vary, most sources cite a temperature range between -1°C and 20°C, with reduced growth and survival when temperatures exceed 18°C. Blue mussels are also typically found in shallower depths (0-24 meters).

```{r, warning = FALSE, message = FALSE}
# Identify suitable habitat for blue mussels
mussel_habitat <- find_suitable_habitat(
  temp_min = -1,
  temp_max = 20,
  depth_min = 0,
  depth_max = 24,
  species_name = "Blue Mussel"
)
```

## **Regional Analysis of Suitable Habitat**

To quantify aquaculture potential by jurisdiction, we calculate the total area of suitable habitat within each state's Exclusive Economic Zone.

### **Calculating Suitable Area by Region**

This function builds upon the habitat suitability analysis by calculating the total area (in square kilometers) of suitable habitat within each region.

```{r, warning = FALSE, message = FALSE}
# Returns a data frame of the area suitable fora species by region
total_habitable_areas <- function(temp_min, temp_max, depth_min, depth_max, species_name){
  
  # Create suitability masks
  sst_suitable <- mean_sst >= temp_min & mean_sst <= temp_max
  depth_suitable <- depth_cropped >= depth_min & depth_cropped <= depth_max
  suitable_habitat <- sst_suitable * depth_suitable
  
  # Rasterize EEZ regions
  eez_raster <- rasterize(eez_raster, suitable_habitat, field = "rgn")
  
  # Calculate cell area 
  cell_area <- cellSize(suitable_habitat, unit = "km")
  
  # Get area of suitable cells in dataframe
  area_suitable <- cell_area * suitable_habitat
  
  # Get the total area that is suitable in each EEZ
  total_area_by_eez <- zonal(area_suitable, eez_raster, fun = "sum", na.rm = TRUE)
  
  return(total_area_by_eez)
}

```

### Table of Suitable Area by Region

```{r}
# Calculate suitable area for each species

oyster_area <- total_habitable_areas(
  temp_min = 11,
  temp_max = 30,
  depth_min = 0,
  depth_max = 70,
  species_name = "Oyster") %>% 
  rename("Oysters" = area)

mussel_area <- total_habitable_areas(  
  temp_min = -1,
  temp_max = 20,
  depth_min = 0,
  depth_max = 24,
  species_name = "Blue Mussel") %>%
  rename("Blue Mussle" = area)


# Join results and create formatted table
habitat_summary <- inner_join(oyster_area, mussel_area, by = "rgn") %>%
  rename("Region" = rgn) %>%
  mutate(across(where(is.numeric), ~round(.x, 2))) %>%
  kbl(
    caption = "Suitable Area For Oysters and Blue Mussles",
    align = c("l", "r", "r")
  ) %>% 
    kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  add_header_above(c(" " = 1, "Suitable Area (km²)" = 2))

habitat_summary

```

**Interpretation**:

1.  **Washington** has the most suitable habitat for blue mussels (260.91 km²), which is more than double all of the other regions. This is likely caused by the cold waters of the state and large amount of shallow water found in regions like the Puget Sound.

2.  **Southern California** leads in oyster habitat suitability (160.53 km²), likely due to their warmer waters.

3.  **Oregon** shows a strong preference for mussels over oysters (76.65 km² vs. 30.11 km²), which is expected to to it's cooler waters.

4.  **Central California** shows relatively low suitability for both species. It is unclear if this is caused by the temperature range or a lack of shallow waters.

### Implications for Aquaculture Development

Although the West Coast appears to have a larger total area suited to blue mussels, oyster farming currently dominates every region. This mismatch between environmental suitability and current industry patterns may indicate **untapped potential for blue mussel aquaculture**, particularly in Washington and Oregon where there is a significantly larger area suitable for mussels than oysters.

### **Limitations**

While this analysis provides a solid foundation for further research, it has some limitations that should be addressed in future research:

1.  **Temporal Resolution**: The data uses annual SST, which does not account for seasonal variations in ocean conditions. Locations which experience extreme seasonal differences are likely to go undetected using this method, which could result in false positives being introduced.

2.  **Environmental factors not considered:** Important variables such as acidity, tidal patterns, wave exposure, and El Nino events were not included in the analysis.

3.  **Socioeconomic and Regulatory Constraints**: Shipping lanes, marine protected areas, and competing ocean uses were not incorporated into the suitability model.

## Future research 

In addition to adressing the above limitation, future research should investigate the economic viability of establishing mussel production in areas identified as suitable. Additionally, climate forecast should be included in the analysis to see if sites are likely to remain suitable in the future.

### **References**

National Oceanic and Atmospheric Administration. (2023). NOAA CoastWatch Pathfinder Sea Surface Temperature Dataset.[**https://podaac.jpl.nasa.gov/dataset/AVHRR_PATHFINDER_L3_V52**](https://podaac.jpl.nasa.gov/dataset/AVHRR_PATHFINDER_L3_V52)

National Centers for Environmental Information. (2023). Gridded Bathymetry Data.[**https://www.ncei.noaa.gov/access/gridded-bathymetry-data**](https://www.ncei.noaa.gov/access/gridded-bathymetry-data)

NOAA Fisheries. (2023). West Coast Regional Boundaries.[**https://www.fisheries.noaa.gov/west-coast/marine-protected-areas/west-coast-regional-boundaries**](https://www.fisheries.noaa.gov/west-coast/marine-protected-areas/west-coast-regional-boundaries)

Food and Agriculture Organization. (2022). The State of World Fisheries and Aquaculture.[**https://www.fao.org/3/ca9229en/ca9229en.pdf**](https://www.fao.org/3/ca9229en/ca9229en.pdf)
